\providecommand{\rootdir}{..}
\documentclass[\rootdir/thesis.tex]{subfiles}

\begin{document}

If the spokes are tensioned beyond a critical value, the circular shape becomes unstable and the rim will buckle into a non-planar shape. The post-buckling configuration is generally stable and the original shape of the wheel can be recovered by uniformly reducing the spoke tensions. Despite its implications for wheel strength, the buckling problem has never received a rigorous treatment. Jobst Brandt alludes to buckling in his practical manual for wheelbuilders\cite{Brandt}:

\begin{quote}
\emph{``If the wheel becomes untrue in two large waves during stress relieving, the maximum, safe tension has been exceeded. Approach this tension carefully to avoid major rim distortions. When the wheel loses alignment from stress relieving, loosen all spokes a half turn before retruing the wheel.''}
\end{quote}

The ``stress relieving'' referred to here is the practice of laterally loading the rim (or the spokes directly) so that the spokes temporarily increase their tension, which is presumed also to cause plastic flow in the spoke elbows so as to lower residual stresses in the service condition, and hence reduces fatigue.

Flexural-torsional buckling of rings can be treated as a special case of buckling of arches, where the included angle is allowed to go to $2\pi$ with appropriately periodic boundary conditions. Timoshenko and Gere\cite{TimoshenkoGere} gave a formula for the critical load for a ring with doubly-symmetric cross-section subjected to a line load directed towards the ring's center The theory of flexural-torsional buckling of monosymmetric arches (bicycle rim sections generally have only one plane of symmetry) was broadly formalized by Trahair and Papangelis\cite{Trahair} using the virtual work approach to derive the equilibrium and stability equations. Their theory has been extended to treat arches with continuous\cite{} or discrete\cite{} elastic supports and elastic end restraints\cite{}.

The problem of the prestressed bicycle wheel is unique for a number of reasons. First, the load for buckling due to spoke over-tension is internal to the structure. Second, the spokes act both as elastic restraints resisting buckling and as prestressing elements causing buckling. Third, the lateral, radial, tangential, and torsional restraining actions of the spokes are commonly coupled – e.g. lateral deflection at a spoke may produce a mix of those reactions on the rim section. These considerations extend to other structural systems. Large observation wheels such as the London Eye\cite{} and the Singapore Flyer\cite{} resemble bicycle wheels and achieve lateral stability due to the bracing angle of prestressed cables, and must be designed against flexural-torsional buckling. At the biological scale, the cellular fragmentation process which leads to platelet formation may also be driven by flexural-torsional buckling instability of a ring of bundled actin fibers loaded by an elastic membrane which both promotes buckling and provides elastic restraint\cite{}.


\section{Elastic stability criterion}
It was previous shown (Section \ref{sec:ModeMatrix}) that the deformed shape of the wheel under external loads could be found by solving the mode matrix equation \eqref{eq:mm_Kd_f}. One form of the elastic stability criterion states that for any admissable deformation, the second variation of the total potential \eqref{eq:mm_TotPot} must be positive. Since any rim deformation must be periodic and continuous, the modes in \eqref{eq:mm_FourierDef} form a set of kinematically admissable deformations. Therefore the buckling tension can be defined as the tension at which \eqref{eq:mm_Kd_f} admits nontrivial solutions when $\mathbf{F}_{ext}=\mathbf{0}.$\todo{Check wording of precise stability criterion}

The spoke stiffness comprises a term proportional to the elastic stiffness, $EA$, of the spokes and a term proportional to the spoke tension, $T$. Additionally, the rim stiffness comprises a term depending on the bending, torsion, and warping constants, and a term proportional to the net radial spoke tension per unit length, $\bar{T}$. In order to make the dependence on $\bar{T}$ explicit, we write the buckling criterion as follows:
\begin{equation}
\label{eq:Tc_crit}
\det{\left[\mathbf{K}_{rim}^{el} + \mathbf{K}_{spokes}^{el} + \bar{T}(\mathbf{K}_{spokes}^{tens} - \mathbf{K}_{rim}^{tens})\right]} = 0
\end{equation}

If the complete details of discrete spokes and possible radial/lateral coupling are retained, a numerical solution to \eqref{eq:Tc_crit} can be obtained by including enough modes such that the relevant length scales (e.g. the distance between spokes) are correctly approximated, and then either numerically solving the characteristic polynomial of \eqref{eq:Tc_crit}, or iteratively increasing $\bar{T}$ until the determinant is minimized. Under more restrictive assumptions, analytical solutions are possible, as shown in Table \ref{tab:BucklingSolutions}.

\begin{table}[h]
\caption{Simplifying assumptions for elastic buckling criterion.}
\label{tab:BucklingSolutions}
\begin{tabularx}{\textwidth}{p{1.2cm}Xccc}
% \hline\noalign{\smallskip}
\toprule
\centering Smeared spokes & \centering Symmetry & Rim & Complexity & Analytical solution\\
\midrule
\centering No  & \centering None & Monosymmetric$^{\rm b}$ & $\det$ of (4+8N)x(4+8N) matrix & None\\
\centering Yes & \centering None & Monosymmetric & $\det$ of 4x4 matrix & Impractical\\
\centering Yes & \centering Left-right$^{\rm a}$& Monosymmetric& Quadratic equation for $\bar{T}_c$ & Solution of \eqref{eq:Tc_quad}\\
\centering Yes & \centering Left-right& Bi-symmetric$^{\rm c}$ & Linear equation for $\bar{T}_c$ & \eqref{...}\\
\centering Yes & \centering Left-right, no spoke offset & Bi-symmetric & Linear equation for $\bar{T}_c$ & \eqref{...}\\
\bottomrule
\multicolumn{5}{p{\textwidth}}{
$^{\rm a}$The distance to the hub flange and spoke arrangement are identical on the left and right sides, which implies no radial/lateral coupling.
}\\
\multicolumn{5}{p{\textwidth}}{
$^{\rm b}$Monosymmetry implies $y_0\neq 0$
}\\
\multicolumn{5}{p{\textwidth}}{
$^{\rm c}$Bi-symmetry implies $y=0$, i.e. the shear center and centroid coincide.
}\\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Smeared spokes approximation}

If the smeared-spokes approximation is used, the mode stiffness matrix $\mathbf{K}_{rim} + \mathbf{\bar{K}}_{spokes}$ has a block diagonal form, with each block corresponding to a different mode $n$ in \eqref{eq:mm_FourierDef}. The first two modes $n=0$ and $n=1$ are rigid-body motions and do not admit buckling. The buckling criterion then becomes

\begin{equation}
\label{eq:Tc_crit_smeared}
\det{\left[\mathbf{K}_{rim,n}^{el} + \mathbf{\bar{K}}_{spokes,n}^{el} + \bar{T}(\mathbf{\bar{K}}_{spokes,n}^{tens} - \mathbf{K}_{rim,n}^{tens})\right]} = 0
\end{equation}

In its most general form, Equation \eqref{eq:Tc_crit} results in a quadratic equation for $\bar{T}$ which can be solved to find the critical tension $\bar{T}_{c,n}$ for a given mode $n$. The critical buckling tension $\bar{T}_c$ for the wheel is the minimum $\bar{T}_{c,n}$ over all integer modes.

The elastic stiffness $EA$ is much less than the spoke tension. Therefore, the tension components of the spoke stiffness matrix $\bar{k}_{ij}^{tens}$ can generally be neglected compared with $\bar{k}_{ij}^{el}$, except for $\kuu$ due to the small lateral projection of the spoke vector. In this section we will neglect the tension component of all stiffness parameters except for $\kuu$.

\subsubsection{No radial/lateral coupling}
If the radial/lateral spoke coupling terms are neglected ($\kuv=\kuw=\kvp=\kwp=0$), the buckling criterion \eqref{eq:Tc_crit} reduces to
\begin{multline}
\label{eq:Tc_quad}
\left(R^2\kuu^{tens}y_0 - n^2Ry_0 \right)\bar{T}^2\\
-\left[\frac{EI_2}{R}\left(n^2 - 3n^4 \frac{y_0}{R} + R\kuu^{tens}\right)
       +\frac{\widetilde{GJ}}{R}\left(n^4 - \frac{y_0}{R}(2n^4 - n^2) - R\kuu^{tens}n^2\right) \right.\\
       \left. +R\kpp (n^2 - R\kuu^{tens}) + 2Ry_0\kup n^2 - R^2\kuu y_0 \vphantom{\frac12} \right]\bar{T}\\
+\left[\frac{EI_2 \widetilde{GJ} n^2}{R^4}(n^2-1)^2
       +EI_2\left(\kuu + 2\frac{\kup}{R}n^2 + \frac{\kpp}{R^2}n^4\right) \right.\\
       \left. +\widetilde{GJ}\left(\kuu n^2 + 2\frac{\kup}{R}n^2 + \frac{\kpp}{R^2}n^2\right)
       +R^2(\kuu\kpp - \kup^2)\right] = 0
\end{multline}

where $\widetilde{GJ} = GJ + EI_wn^2/R^2$ is the effective torsional stiffness.

\subsubsection{No radial/lateral coupling, bi-symmetric rim}
The quadratic term in \eqref{eq:Tc_quad} is proportional to $y_0$. If the rim cross-section is assumed to be symmetric across both the $\eo$ and $\et$ axes, then the buckling criterion \eqref{eq:Tc_quad} reduces to a linear equation for $\bar{T}$. Using the non-dimensional paremeters defined in Section \ref{sec:Lateral}, the non-dimensionalized critical buckling tension is given by
\begin{equation}
\label{eq:Tcn_lin}
t_{c,n} = \left(\frac{1}{n^2 - R\kuu^{tens}}\right)
\left(\lambda_{uu}
      +\frac{(n^4 + \mu n^2)\lambda_{\phi\phi}
             +2n^2(\mu + 1)\lambda_{u\phi} - \lambda_{u\phi}^2
             +\mu n^2(n^2-1)^2}
        {1 + \mu n^2 + \lambda_{\phi\phi}}\right)
\end{equation}

The first term $(n^2 - R\kuu^{tens})^{-1}$ accounts for the fact the spokes (and therefore the direction of the applied tension at the rim) rotates under a buckling displacement. In classical buckling analysis, this accounts for the difference between buckling under dead loads (e.g. gravity) and directed loads (e.g. tensioned cables).

\subsubsection{No radial/lateral coupling, bi-symmetric rim, no spoke offset}
If the spokes are assumed to connect to the rim through the shear center, the lateral/torsional coupling terms vanish, i.e. $\lambda_{u\pi}=\lambda_{\phi\phi}=0$. Thus, the critical buckling tension is given by

\begin{equation}
\label{eq:Tcn_lin_nophi}
t_{c,n} = \left(\frac{1}{n^2 - R\kuu^{tens}}\right)
\left(\lambda_{uu}
      +\frac{\mu n^2(n^2-1)^2}
        {1 + \mu n^2}\right)
\end{equation}

\todo{Equivalent spring model}

\section{Limiting solutions of \eqref{eq:Tcn_lin_nophi} for special cases}

\section{Buckling of asymmetric wheels}

\section{Finite-element buckling calculations}

\end{document}
