\providecommand{\rootdir}{..}
\documentclass[\rootdir/thesis.tex]{subfiles}

\begin{document}

\section{Kinematics and strain energy}
A schematic of a typical bicycle wheel is shown in Figure \ref{fig}. The structure consists of a hub, rim, and spokes. The spokes are connected to two parallel flanges on the hub and the resulting projected bracing angle, $\alpha$,  stabilizes the rim laterally. In this way, the spokes are analogous to guy-wires. Conventional spokes are threaded into nipples set into the rim which can be tightened and loosened independently. The spokes are tensioned during construction to prevent them from buckling when the wheel carries load.

Bicycle rims today are typically constructed from thin-walled extruded sections, while rims on older bicycles were constructed from roll-formed metal strip, or sometimes solid wood. To describe forces and deformations of a wheel, we employ a local coordinate triad whose origin is at the cross section shear center, $S$. (The shear center is that unique point in a beam cross section at which a shear load causes no twist.) The basis vector $\mathbf{e}_2$ is radially inwards, the basis vector $\mathbf{e}_3$ is in the circumferential direction of increasing arc length $s$, and the lateral basis vector $\mathbf{e}_1$ completes a right-hand triad.

\subsection{Deformation of the rim}

The rim is modeled as a circular beam of constant cross-section with an axis of symmetry in the plane of the wheel. The height of the shear center $s$ relative to the centroid is $y_0$. The displacement of the rim is described in terms of the cross-section displacement $u\mathbf{e}_1 + v\mathbf{e}_2 + w\mathbf{e}_3$ in polar coordinates and rotation angle $\phi$ of the cross-section about $\mathbf{e}_3$.

The deformation of the rim is then defined by the change in curvatures and twist of its line of shear centers, and longitudinal extension of its centerline (which differs from the line of shear centers if the centroid is offset). To compute these quantities, we begin with the rotation of each cross-section, defined as a vector in terms of the cylindrical basis vectors. According to Euler-Bernoulli theory, the cross sections do not shear relative to the beam axis, so rotation is defined by slopes relative to the axis, plus a twisting rotation. Thus the cross section rotation vector is $(v + w/R) \mathbf{e}_1 + u' \mathbf{e}_2 + \phi\mathbf{e}_3$, where $()'$ denotes the derivative with respect to circumferential arc length $s$. We differentiate with respect to $s$ to get a bending and twist vector, recalling that $\et' = -\eh/R$ and $\eh' = \et/R$. The result is:
\begin{subequations}
\label{eq:kappa}
\begin{align}
\mathbf{\kappa} = \kappa_1\eo + \kappa_2\et + \kappa_3\eh\\
\kappa_1 = \left( v'' + \frac{w'}{R} \right) \label{eq:kappa1}\\
\kappa_2 = \left( u'' + \frac{\phi}{R} \right) \label{eq:kappa2}\\
\kappa_3 = \left( \phi' - \frac{u'}{R} \right) \label{eq:kappa3}
\end{align}
\end{subequations}

We define these three components as in-plane bending, out-of-plane bending, and twisting. For understanding rim deformation, it is worth remarking that $\phi=$ constant creates pure bending (ring eversion), while $u'=$ constant creates pure torsion (analogous to a helical spring).

To determine strain of the centroidal line, we need the displacement vector of the centroid: $\mathbf{u}_c = (u+y_0\phi)\eo + v\et + (w+y_0 (v'+w/R))\eh$. We differentiate with respect to $s$ to determine the $\eh$ component or extensional strain:
\begin{equation}
\label{eq:mem_strain}
\varepsilon_m = w' - \frac{v}{R} + y_0\left(v'' + \frac{w'}{R} \right)
\end{equation}

Due to the prestressing process, the reference configuration is not a state of zero stress. In addition to the tension in the spokes, the rim is under compression\cite{Sharp} (proportional to the tension in the spokes) and may also support a uniform radial bending moment about $\eo$ depending on the construction method of the rim, as well as a periodic bending moment from the separation of discrete spokes. Most modern rims are constructed from profiles permanently deformed into a circle close to its final radius, so the uniform moment can generally be neglected. We will show that the periodic moment can be neglected provided the wheel has a sufficient number of spokes compared with its in-plane bending stiffness.

The increase in strain energy in the rim in moving from the reference configuration to a deformed configuration can be decomposed into components for centroidal axial stretching, radial and lateral bending, twisting, and varied warping due to twist gradient:
\begin{equation}
\label{eq:U_rim}
U_{rim} = \frac{1}{2} \int_0^{2\pi R}[EA\varepsilon_m^2 + EI_1\kappa_1^2+EI_2\kappa_2^2 + GJ\kappa_3^2 + EI_w(\kappa_3')^2]ds
\end{equation}

The warping energy $EI_w(\kappa_3')^2$ is primarily related to the bending energy in the rim sidewalls (which deform in a similar manner as the flanges of an I-beam) of the rim due to varying torsion. For a single-wall rim, the resistance to warping can account for most of the effective torsional stiffness.

\subsection{Deformation of the spoke system}

\subfile{\rootdir/chapters/stress_analysis/spoke_stiffness}


\section{Stress and deformation of the pretensioned wheel}

\subfile{\rootdir/chapters/stress_analysis/radial_bulging}


\section{Stresses in the wheel under external loads}

We assume the rim is loaded by distributed forces $f_u,f_v,f_w$ and a distributed moment $m$. The total potential energy in the deformed configuration under this system of loads is
\begin{equation}
\label{eq:TotPot}
\Pi = U_{rim} + U_{spokes} - (f_uu+f_vv+f_ww+m\phi)
\end{equation}

The solution for the displacements is found by minimizing \eqref{eq:TotPot}. The Euler-Lagrange equations for \eqref{eq:TotPot} give a set of four coupled differential equations for the displacement variables:
\begin{equation}
\label{eq:EulerLagrange}
\begin{bmatrix}
 \D_{vv} + \kvv & \D_{vw} + \kvw & \kuv              & \kvp\\
-\D_{vw} + \kvw & \D_{ww} + \kww & \kuw              & \kwp\\
\kuv            & \kuw           & \D_{uu} + \kuu    & \D_{u\phi} + \kup\\
\kvp            & \kwp           & \D_{u\phi} + \kup & \D_{\phi\phi} + \kpp
\end{bmatrix}
\begin{bmatrix}
v\\w\\u\\\phi
\end{bmatrix}=
\begin{bmatrix}
f_v\\f_w\\f_u\\m
\end{bmatrix}
\end{equation}

where
\begin{align*}
\D_{vv} &= (EI_1 + EAy_0^2)\D^4 - 2EA\left(\frac{y_0}{R}\right)\D^2 + \frac{EA}{R^2}\\
\D_{ww} &= -\left(\frac{EI_1}{R^2} + EA\left(1+\frac{y_0}{R}\right)^2\right)\D^2\\
\D_{vw} &= \left(\frac{EI_1}{R} + EAy_0\left(1+\frac{y_0}{R}\right)\right)\D^3 - \frac{EA}{R}\left(1+\frac{y_0}{R}\right)\D\\
\D_{uu} &= \left(EA_2 + \frac{EI_w}{R^2}\right)\D^4 - \frac{GJ}{R^2}\D^2\\
\D {u\phi} &= \frac{EI_w}{R}\D^4 - \left(\frac{EI_2}{R} + \frac{GJ}{R}\right)\D^2\\
\D_{\phi\phi} &= EI_w\D^4 - GJ\,\D^2 + \frac{EI_2}{R^2}
\end{align*}

and $\D^n \equiv (d/ds)^n$.

Due to the spoke stiffness parameters $\bar{k}_{ij}$, all of the displacement variables appear in each of the equilibrium equations. In the most general case with no further simplifications or symmetries, the combined governing equation is a 14\textsuperscript{th} order linear ordinary differential equation with constant coefficients. In most cases relevant to real wheels many of the coupling terms $\bar{k}_{ij}$ $(i \neq j)$ are identically zero or are small compared to other relevant quantities. For example, in a wheel with either mirror symmetry or mirror-rotational symmetry across the plane of the wheel, all the off-diagonal terms $\bar{k}_{ij}$ $(i \neq j)$ are identically zero, except $\bar{k}_{u\phi}$. In the case where $\bar{k}_{uv}=\bar{k}_{uw}=\bar{k}_{v\phi}=\bar{k}_{w\phi}=0$, the equilibrium equations decouple into a pair of equations for radial-tangential deformations and a pair of equations for lateral-torsional deformations.

\subsection{Loads in the plane of the wheel}

\subfile{\rootdir/chapters/stress_analysis/radial_tangential}

\subsection{Loads out of the plane of the wheel}

Under lateral loads, the rim bends and twists into a non-planar shape. The wheel is considerably more flexible in the lateral direction than in the radial direction due to the small lateral projection of the spokes. When the rim undergoes lateral deformation, the potential energy of the compressive load induced by the spoke pretension is reduced. Due to the large lateral compliance, this reduction in potential energy can be significant compared to the increase in strain energy due to lateral bending and twisting. This leads to larger lateral deflections than would be predicted by the linear theory in Eqn. \eqref{eq:EulerLagrange}, with the possibility of lateral-torsional instability at a sufficiently high spoke tension.

We consider a variation of the displacements from the prestressed reference configuration defined by $(u,v,w,\phi)=(0,v_0,w_0,0)$ to a new deformed state defined by $(u,v,w,\phi)=(0+u,v_0,w_0,0+\phi)$. Trahair and Papangelis\cite{Trahair} give the virtual work due to a uniform compressive stress in a monosymmetric curved beam, assuming that the displacements $v_0,w_0$ are small compared to $u,r\phi$
\begin{equation}
\label{eq:V_rim}
V_{rim} = \frac{1}{2}R\bar{T} \int_0^{2\pi R}
    \left[u'^2 + r_x^2\phi'^2+r_y^2\left(\frac{u'}{R}-\phi'\right)^2
          +y_0\left(2u''\phi-\frac{\phi^2}{R}\right) + y_0^2\phi'^2 \right] ds
\end{equation}

Then the total potential energy, Equation \ref{eq:TotPot} is augmented by the virtual work due to compressive stress in the rim:
\begin{equation}
\label{eq:TotPot_V}
\Pi = U_{rim} + U_{spokes} - V_{rim} - (f_uu+m\phi)
\end{equation}

The Euler-Lagrange equations that guarantee the minimization of \eqref{eq:TotPot_V} are
\todo{Check full lateral-torsional equations with y0, rx, ry}
\begin{subequations}
\label{eq:lat_ode_2}
\begin{align}
\begin{split}
  EI_2\left( \ds{u}{4} - \frac{1}{R}\ds{\phi}{2} \right)
  + \frac{EI_w}{R^2}\left( R\ds{\phi}{4} + \ds{u}{4} \right)
  - \frac{GJ}{R^2}\left( R\ds{\phi}{2} + \ds{u}{2} \right)\\
  - R\bar{T}\left( \ds{u}{2} - y_0\ds{\phi}{3}
                  -\frac{r_y^2}{R^2}\left(R\ds{\phi}{2}
                  -\frac{1}{R}\ds{u}{2}\right) \right)
  + \kuu u + \kup \phi = f_u
\end{split}\\
\begin{split}
  - \frac{EI_2}{R}\left( \ds{u}{2} - \frac{1}{R}\phi \right)
  + EI_w\left( \ds{\phi}{4} + \frac{1}{R}\ds{u}{4} \right)
  - GJ\left( \ds{\phi}{2} + \frac{1}{R}\ds{u}{2} \right)\\
  + R\bar{T}\left((r_x^2+r_y^2+y_0^2)\ds{\phi}{2}
                  -\left(\frac{r_y^2}{R}\right)\ds{u}{2}
                  -\left(\frac{y_0}{R}\right)\phi\right)
  + \kpp \phi + \kup u = m
\end{split}
\end{align}
\end{subequations}

This coupled system leads to an eighth-order ordinary differential equation with constant coefficients and all even order derivatives. Analytical formulae for the roots are possible, but impractical due to the need to find roots of a quartic characteristic polynomial. Neglecting warping stiffness and defining the non-dimensional groups $\lambda_{uu}=\kuu R^4/EI_2$, $\mu=GJ/EI_2$, and $t=R^4\bar{T}/EI_2$, Equations \eqref{eq:lat_ode_2} become:
\begin{multline}
\label{eq:lat_ode_full}
  \left(1-\frac{t}{\mu}\frac{r_0^2}{R^2} - \frac{t^2}{\mu}\frac{y_0^2}{R^2}\right)\dt{u}{6}\\
  +\left(2+t-\frac{t}{\mu}\frac{y_0}{R} + 2\frac{t}{\mu}\frac{r_y^2}{R^2}
         +t\left(\frac{r_0^2 + 3r_y^2}{R^2}\right)
         -\frac{t^2}{\mu}\left(\frac{r_0^2}{R^2}+\frac{r_x^2r_y^2+r_y^2y_0^2}{R^4}\right)
         +\frac{\lambda_{\phi\phi}}{\mu}\right) \dt{u}{4}\\
  +\left(1+\lambda_{uu}+\lambda_{\phi\phi}-2\lambda_{u\phi}
         -\frac{t}{\mu}+t\frac{y_0}{R}-t\frac{\lambda_{uu}}{\mu}\frac{r_0^2}{R^2}
         -t\frac{\lambda_{\phi\phi}}{\mu}
         -\frac{t^2}{\mu}\left(\frac{y_0^2}{R^2}-(2\lambda_{u\phi}+\lambda_{\phi\phi})\frac{r_y^2}{R^2} - \frac{r_y^2y_0}{R^3}\right)
         \right)\dt{u}{2}\\
  -\frac{1}{\mu}\left(\lambda_{uu}\left(1+t\frac{y_0}{R}\right)
                      +\lambda_{uu}\lambda_{\phi\phi} + \lambda_{u\phi}^2\right)u
\end{multline}

where $r_0^2=r_x^2+r_y^2+y_0^2$. Neglecting second-order quantities $r_x^2,r_y^2,y_0^2$ compared to $R^2$, Equation \eqref{eq:lat_ode_full} becomes:
\begin{multline}
\label{eq:lat_ode}
  \dt{u}{6}
  +\left(2+t-\frac{t}{\mu}\frac{y_0}{R}
         +\frac{\lambda_{\phi\phi}}{\mu}\right) \dt{u}{4}
  +\left(1+\lambda_{uu}+\lambda_{\phi\phi}-2\lambda_{u\phi}
         -\frac{t}{\mu}+t\frac{y_0}{R}
         -t\frac{\lambda_{\phi\phi}}{\mu}\right)\dt{u}{2}\\
  -\frac{\lambda_{uu}}{\mu}\left(1+\lambda_{\phi\phi}
                                 -\frac{\lambda_{u\phi}^2}{\lambda_{uu}}
                                 +t\frac{y_0}{R}\right)u
\end{multline}

Dropping the terms in Equation \eqref{eq:lat_ode} involving $t, y_0, \lambda_{u\phi}, \lambda_{\phi\phi}$, one recovers the differential equation derived by Pippard\cite{Pippard}. The dimensionless tension $t$ can be interpreted as the spoke tension divided by the spoke tension which would create a compressive stress in the rim equal to the buckling load of a straight fixed-fixed column of length $2\pi R$. The stiffness ratio $\lambda_{uu}=\lambda_{el}+\lambda_t$ is composed of two parts: $\lambda_{el}$, a term proportional to the elastic stiffness of the spokes and $\lambda_t$, a term proportional to the spoke tension. For most bicycle wheels, $\lambda_t$ can be obtained to a very good approximation as $\lambda_T = R^2l_s\bar{T}/EI_2 \approx t$.

\section{Limits of the continuum approximation}


\section{Mode stiffness matrix method}

\end{document}
