\providecommand{\rootdir}{../..}
\documentclass[\rootdir/thesis.tex]{subfiles}

\begin{document}

Under lateral loads, the rim bends and twists into a non-planar shape. The wheel is considerably more flexible in the lateral direction than in the radial direction due to the small lateral projection of the spokes. When the rim undergoes lateral deformation, the potential energy of the compressive load induced by the spoke pretension is reduced. Due to the large lateral compliance, this reduction in potential energy can be significant compared to the increase in strain energy due to lateral bending and twisting. This leads to larger lateral deflections than would be predicted by the linear theory in Eqn. \eqref{eq:EulerLagrange}, with the possibility of lateral-torsional instability at a sufficiently high spoke tension.

We consider a variation of the displacements from the prestressed reference configuration defined by $(u,v,w,\phi)=(0,v_0,w_0,0)$ to a new deformed state defined by $(u,v,w,\phi)=(0+u,v_0,w_0,0+\phi)$. Trahair and Papangelis\cite{Trahair} give the virtual work due to a uniform compressive stress in a monosymmetric curved beam, assuming that the displacements $v_0,w_0$ are small compared to $u,r\phi$
\begin{equation}
\label{eq:V_rim}
V_{rim} = \frac{1}{2}R\bar{T} \int_0^{2\pi R}
    \left[u'^2 + r_x^2\phi'^2+r_y^2\left(\frac{u'}{R}-\phi'\right)^2
          +y_0\left(2u''\phi-\frac{\phi^2}{R}\right) + y_0^2\phi'^2 \right] ds
\end{equation}

Then the total potential energy, Equation \ref{eq:TotPot} is augmented by the virtual work due to compressive stress in the rim:
\begin{equation}
\label{eq:TotPot_V}
\Pi = U_{rim} + U_{spokes} - V_{rim} - (f_uu+m\phi)
\end{equation}

The Euler-Lagrange equations that guarantee the minimization of \eqref{eq:TotPot_V} are
\todo{Check full lateral-torsional equations with y0, rx, ry}
\begin{subequations}
\label{eq:lat_ode_2}
\begin{align}
\begin{split}
  EI_2\left( \ds{u}{4} - \frac{1}{R}\ds{\phi}{2} \right)
  + \frac{EI_w}{R^2}\left( R\ds{\phi}{4} + \ds{u}{4} \right)
  - \frac{GJ}{R^2}\left( R\ds{\phi}{2} + \ds{u}{2} \right)\\
  - R\bar{T}\left( \ds{u}{2} - y_0\ds{\phi}{3}
                  -\frac{r_y^2}{R^2}\left(R\ds{\phi}{2}
                  -\frac{1}{R}\ds{u}{2}\right) \right)
  + \kuu u + \kup \phi = f_u
\end{split}\\
\begin{split}
  - \frac{EI_2}{R}\left( \ds{u}{2} - \frac{1}{R}\phi \right)
  + EI_w\left( \ds{\phi}{4} + \frac{1}{R}\ds{u}{4} \right)
  - GJ\left( \ds{\phi}{2} + \frac{1}{R}\ds{u}{2} \right)\\
  + R\bar{T}\left(r_0^2\ds{\phi}{2}
                  -\left(\frac{r_y^2}{R}\right)\ds{u}{2}
                  -\left(\frac{y_0}{R}\right)\phi\right)
  + \kpp \phi + \kup u = m
\end{split}
\end{align}
\end{subequations}

where $r_0^2 = r_x^2+r_y^2+y_0^2$. This coupled system leads to an eighth-order ordinary differential equation with constant coefficients and all even order derivatives. Analytical formulae for the roots are possible, but impractical due to the need to find roots of a quartic characteristic polynomial. Neglecting warping stiffness and defining the non-dimensional groups $\lambda_{uu}=\kuu R^4/EI_2$, $\mu=GJ/EI_2$, and $t=R^4\bar{T}/EI_2$, Equations \eqref{eq:lat_ode_2} become:\todo{Check against Sympy}
\begin{multline}
\label{eq:lat_ode_full}
  \left(1-\frac{t}{\mu}\frac{r_0^2}{R^2} - \frac{t^2}{\mu}\frac{y_0^2}{R^2}\right)\dt{u}{6}\\
  +\left(2+t-\frac{t}{\mu}\frac{y_0}{R} + 2\frac{t}{\mu}\frac{r_y^2}{R^2}
         +t\left(\frac{r_0^2 + 3r_y^2}{R^2}\right)
         -\frac{t^2}{\mu}\left(\frac{r_0^2}{R^2}+\frac{r_x^2r_y^2+r_y^2y_0^2}{R^4}\right)
         +\frac{\lambda_{\phi\phi}}{\mu}\right) \dt{u}{4}\\
  +\left(1+\lambda_{uu}+\lambda_{\phi\phi}-2\lambda_{u\phi}
         -\frac{t}{\mu}+t\frac{y_0}{R}-t\frac{\lambda_{uu}}{\mu}\frac{r_0^2}{R^2}
         -t\frac{\lambda_{\phi\phi}}{\mu}
         -\frac{t^2}{\mu}\left(\frac{y_0^2}{R^2}-(2\lambda_{u\phi}+\lambda_{\phi\phi})\frac{r_y^2}{R^2} - \frac{r_y^2y_0}{R^3}\right)
         \right)\dt{u}{2}\\
  -\frac{1}{\mu}\left(\lambda_{uu}\left(1+t\frac{y_0}{R}\right)
                      +\lambda_{uu}\lambda_{\phi\phi} + \lambda_{u\phi}^2\right)u = 0
\end{multline}

Neglecting second-order quantities $r_x^2,r_y^2,y_0^2$ compared to $R^2$, Equation \eqref{eq:lat_ode_full} becomes:\todo{Check against Sympy}
\begin{multline}
\label{eq:lat_ode}
  \dt{u}{6}
  +\left(2 + t - \frac{t}{\mu}\frac{y_0}{R}
         +\frac{\lambda_{\phi\phi}}{\mu}\right) \dt{u}{4}\\
  +\left(1 + \lambda_{uu} + \left(1-\frac{t}{\mu}\right)\lambda_{\phi\phi}
         -2\left(1 + \frac{1}{\mu}\right)\lambda_{u\phi}
         -\frac{t}{\mu}+t\frac{y_0}{R}
         -\frac{t^2}{\mu}\frac{y_0}{R}\right)\dt{u}{2}\\
  -\frac{\lambda_{uu}}{\mu}\left(1+\lambda_{\phi\phi}
                                 -\frac{\lambda_{u\phi}^2}{\lambda_{uu}}
                                 +t\frac{y_0}{R}\right)u = 0
\end{multline}

Dropping the terms in Equation \eqref{eq:lat_ode} involving $t, y_0, \lambda_{u\phi}, \lambda_{\phi\phi}$, one recovers the differential equation derived by Pippard\cite{Pippard}. The dimensionless tension $t$ can be interpreted as the spoke tension divided by the spoke tension which would create a compressive stress in the rim equal to the buckling load of a straight fixed-fixed column of length $2\pi R$. The stiffness ratio $\lambda_{uu}=\lambda_{el}+\lambda_t$ is composed of two parts: $\lambda_{el}$, a term proportional to the elastic stiffness of the spokes and $\lambda_t$, a term proportional to the spoke tension. For most bicycle wheels, $\lambda_t$ can be obtained to a very good approximation as $\lambda_T = R^2l_s\bar{T}/EI_2 \approx t$.

\subsubsection{Solutions to \eqref{eq:lat_ode_2} by the mode stiffness method.}

Equation \eqref{eq:lat_ode} is a 6\textsuperscript{th}-order linear ordinary differential equation with constant coefficients. Since all the derivatives have even order, the roots of the characteristic equation can be solved analytically by solving a cubic equation. However, a straightforward approximation of arbitrary accuracy which preserves warping stiffness is possible using the Rayleigh-Ritz method.

Restricting our analysis to loading which is symmetric about a point $\theta=0$, we assume that the displacements take the form $u=\sum u_n \cos{n\theta}$, $\phi=\sum \phi_n \cos{n\theta}$. With this basis, the total potential energy \eqref{eq:TotPot_V} takes the form
\begin{equation}
\label{eq:TotPot_RR}
\Pi = \sum_{n=0}^{N} U_n - \sum_{n=0}^{N} V_n - P\sum_{n=0}^{N} u_n
\end{equation}

where $U_n$ and $V_n$ are the strain energy and the virtual work due to rim compression for the $n$\textsuperscript{th} mode, and $P$ is the applied load. The strain energy and virtual work decompose additively due to the orthogonality of the Fourier basis functions around the unit circle. The mode coefficients $u_n, \phi_n$ are those which bring Equation \eqref{eq:TotPot_RR} to a global minimum. Since \eqref{eq:TotPot_RR} is quadratic in $u_n, \phi_n$ it is sufficient to require
\begin{subequations}
\label{eq:RR_min}
\begin{align}
\frac{\partial \Pi}{\partial u_n} &= 0\\
\frac{\partial \Pi}{\partial \phi_n} &= 0
\end{align}
\end{subequations}

% \begin{equation}
% \begin{cases}
% \begin{matrix}
% 2\pi R \kuu u_0 + 2\pi R \kup \phi_0=P\\
% \kup u_0 + \left(\kpp + \frac{EI_2}{R}\right)\phi_0=0
% \end{matrix}, & (n=0)\\
% \begin{matrix}
% \left(\frac{EI_2\pi}{R^3}n^4 + \frac{\tilde{GJ}\pi}{R^3}n^2 + \pi R\kuu -\frac{n^2\bar{T}}{2}\right)u_n
% 	-\left(\frac{EI_2\pi}{R^2}n^2 + \frac{\tilde{GJ}\pi}{R^2}n^2-\pi R\kup\right)\phi_n =P\\
% \left(\frac{EI_2\pi}{R^2}n^2 + \frac{\tilde{GJ}\pi}{R^2}n^2 - \pi R\kup \right)u_n
%  - \left(\frac{EI_2\pi}{R} + \frac{\tilde{GJ}\pi}{R}n^2 + \pi R\kpp\right)\phi_n =0
% \end{matrix}, & (n\geq 1)
% \end{cases}
% \end{equation}

The solution to \eqref{eq:RR_min} for $u_n,\phi_n$ is



\end{document}